---
title: "Bellabeat"
author: "meiyin"
date: "2022-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Urška Sršen and Sando Mur founded Bellabeat, a high-tech company that manufactures health-focused smart products.
Sršen used her background as an artist to develop beautifully designed technology that informs and inspires women around
the world. Collecting data on activity, sleep, stress, and reproductive health has allowed Bellabeat to empower women with knowledge about their own health and habits. Since it was founded in 2013, Bellabeat has grown rapidly and quickly
positioned itself as a tech-driven wellness company for women.

## Step 1: Ask
## Task
To analyse non-Bellabeat smart devices and its usage by consumers in order to apply these insights on Bellabeat products to drive sales.

### Key stakeholders
*Urška Sršen: Bellabeat's cofounder and Chief Creative Officer
*Sando Mur: Mathematician and Bellabeat’s co-founder

### Questions to explore:
1. What are some trends in smart device usage?
2. How could these trends apply to Bellabeat customers?
3. How could these trends help influence Bellabeat marketing strategy?

## Step 2: Prepare
## About dataset 
The dataset used in this case study is from Kaggle: [Fitbit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit), a CC0: Public Domain, dataset that is available through [Mobius](https://www.kaggle.com/arashnic).This Kaggle data set contains personal fitness tracker from thirty fitbit users. Thirty eligible Fitbit users consented to the submission of personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring. It includes information about daily activity, steps, and heart rate that can be used to explore users’ habits.

This dataset is downloaded and uploaded to RStudio. It was then analyse using the ROCC approach:

* Reliability: LOW. Dataset is collected from 30 Fitbit users but their gender is unknown. 

* Originality: LOW. Third party collected the data; Amazon Mechanical Turk.

* Comprehensive: MEDIUM. It contains a number of information such as sleep activity, daily activity and their weight. However, it does not contain information on one's gender, age and their health status. 

* Current: LOW. This dataset was collected in 2016 and it is considered old as people's fitness level may change over time.

* Cited: LOW. It does state that the dataset was from Amazon Mechanical Turk; however, there is no link or information regarding this.

### Install and load packages

```{r}
#install.packages(c("tidyverse", "lubridate", "tidyr", "skimr", "janitor", "dplyr", "ggplot2"))
```

```{r}
#lapply(c("tidyverse", "lmtest", "skimr", "janitor", "lubridate", "tidyr", "dplyr", "ggplot2"), require, character.only = TRUE)

```

```{r}
library(tidyverse)
library(lubridate)
library(tidyr)
library(skimr)
library(janitor)
library(dplyr)
library(ggplot2)
```


# Step 3: Process

## Importing datasets
```{r Daily Activity}
activity<- read.csv("archive/Fitabase/dailyActivity_merged.csv")
```

```{r Sleep}
sleep<- read.csv("archive/Fitabase/sleepDay_merged.csv")
```

```{r Weight}
weight<- read.csv("archive/Fitabase/weightLogInfo_merged.csv")
```

## Clean and process data
```{r}
head(activity)
head(weight)
head(sleep)
```

```{r Check the datatype of each column}
str(activity)
str(weight)
str(sleep)
```
It seems that the datatype of ActivityDate in both activity and calories dataframe is character instead of Datetime. So we need to change it. Same goes to the sleep and weight dataframe where the "Date" and "SleepDay" has both date and time and the datatype is character as well. 
```{r Change datatype from character to datetime for activity}
activity$ActivityDate<-as.Date(activity$ActivityDate, format="%m/%d/%Y")
activity$Date<- as.Date(activity$ActivityDate, format="%m/%d/%Y")
```

```{r Convert Date from weight dataset from character and split to date and time}
weight$Date<-as.POSIXct(weight$Date, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
weight$Date_2 <- as.Date(weight$Date, format = "%m/%d/%Y")
weight$Time <- format(as.POSIXct(weight$Date), format = "%H:%M:%S")
weight$Month <- format(weight$Date, format = "%B")

```

```{r Convert SleepDay from sleep dataset from character and split to date and time}
sleep$Date<-as.Date(sleep$SleepDay, format="%m/%d/%Y")
sleep$SleepDay<-as.POSIXct(sleep$SleepDay, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
```

```{r Check for null values : activity dataset}
apply(activity, 2, function(x) any(is.na(x)))
```
There are null values for ActivityDate.

```{r Check for null values : sleep dataset}
apply(sleep, 2, function(x) any(is.na(x)))
```
There are no null values for sleep dataset.

```{r Check for null values : weight dataset}
apply(weight, 2, function(x) any(is.na(x)))
```
There are null values for weight dataset which is Fat.

### Clean the columns accordingly
```{r}
activity<-clean_names(activity)
``` 

```{r sleep}
sleep <- clean_names(sleep)
```

```{r weight}
weight <- clean_names(weight) %>%
  na.omit(weight)
```

### Check again to see if there are any null values left
```{r weight dataset}
apply(weight, 2, function(x) any(is.na(x)))
```
# Step 4: Analyse
## Explore data
```{r Count unique values in each dataset}
n_distinct(activity$id)
n_distinct(sleep$id)
n_distinct(weight$id)
```
There are 33 participants in the activity dataset, 24 participants in the sleep datasets and 2 participants in the weight dataset. Only 2 participants in the weight dataset; therefore, more data is required in order to make any recommendations or conclusions that is concrete and this dataset will not be used for further analysis.

```{r}
# activity
activity %>%
  select(total_steps, total_distance, sedentary_minutes, calories) %>%
summary()


# explore active distance per minutes in activity
activity %>%
  select(very_active_minutes, fairly_active_minutes, lightly_active_minutes) %>%
summary()


# sleep
sleep %>%
  select(total_sleep_records, total_minutes_asleep, total_time_in_bed) %>%
summary()


# weight
weight %>%
  select(weight_kg, bmi) %>%
summary()
```
### Summary
* The average sedentary time is 991.2 minutes which is around 16.5 hours. This has to be reduced as it can have a number of negative effects to the health such as obesity [Mayo Clinic-Risks of being sedentary](https://www.mayoclinic.org/healthy-lifestyle/adult-health/expert-answers/sitting/faq-20058005)

* Majority of the participants fall under the lightly active group.

* The average number of steps per day is 7638. According to CDC, the recommended steps per day is 10,000.

* On average, people sleep for 7 hours.

## Merge data

```{r}
activity_sleep <- merge(activity, sleep, by= c("id", "date"))
head(activity_sleep)
```


# Step 5: Share
## Visualisation + Exploration
### Total Steps vs Calories
```{r}
ggplot(data = activity) + aes(x = total_steps, y=calories) + labs(title = "Total Steps vs Calories") + geom_point(color = "blue") 
```
```{r}
cor(activity$total_steps, activity$calories)
```
There is a moderate positive correlation between the number of steps and calories. This indicates that the more steps the participants take, the more calories they burn. 

### Time Asleep vs Time in Bed
```{r}
ggplot(data = sleep) + aes(x = total_time_in_bed, y=total_minutes_asleep) + labs(title = "Time Asleep vs Time in Bed") + geom_point(color = "salmon") 
```
```{r}
cor(sleep$total_time_in_bed, sleep$total_minutes_asleep)
```
We can clearly see that there is a high positive correlation between time spent in bed and and time spent sleeping. 

### Sleep Duration vs Sedentary Time
```{r}
ggplot(data = activity_sleep) + aes(x = sedentary_minutes, y=total_minutes_asleep) + labs(title = "Sleep Duration vs Sedentary Time") + geom_point(color = "purple") 
```
```{r Correlation}
cor(activity_sleep$sedentary_minutes, activity_sleep$total_minutes_asleep)
```
There is a moderate negative correlation between sleep duration and sedentary time. This means that when a person is less active, they will have a shorter sleep duration.

### Which day are participants more active?
```{r}
activity_sleep <- mutate(activity_sleep, 
                    day = wday(date, label = TRUE))

summarised_activity_sleep <- activity_sleep %>%
  group_by(day) %>%
  summarise(AvgDailySteps = mean(total_steps),
            AvgAsleepMinutes = mean(total_minutes_asleep),
            AvgAwakeTimeInBed = mean(total_time_in_bed), 
            AvgSedentaryMinutes = mean(sedentary_minutes),
            AvgLightlyActiveMinutes = mean(lightly_active_minutes),
            AvgFairlyActiveMinutes = mean(fairly_active_minutes),
            AvgVeryActiveMinutes = mean(very_active_minutes), 
            AvgCalories = mean(calories))
                                    
head(summarised_activity_sleep)
  
```
```{r}
ggplot(data = summarised_activity_sleep) + aes(x = day, y = AvgDailySteps, fill = day) + geom_bar(stat = "identity") + labs(title = "Daily Steps Taken")
```
We can see that most people are active on Saturday and least active on Sunday. 

### Which days are participants sleeping the most?
```{r}
ggplot(data = summarised_activity_sleep) + aes(x = day, y = AvgAsleepMinutes, fill = day) + geom_bar(stat = "identity") + labs(title = "Daily Sleep Duration")
```
Most people sleep more on Sundays. 

## Summaries + Recommendations
* The average number of steps people take per day is 7638. According to CDC, people who take at least 8000 steps per day had shown a 50% reduction of risk of dying [Link to research done by CDC](https://www.nih.gov/news-events/nih-research-matters/number-steps-day-more-important-step-intensity). Therefore, Bellabeat can implement a feature where there is a section in the app that educates people on the benefits of it and also encouraging their users to at take at least 8000 steps per day.

* As the majority of users are under the lightly active category, Bellabeat can have a progression system where users have to reach a certain number of steps or certain number of workouts in order to collect points. Perhaps when Bellabeat has the funds or capacity to allow users to change the points collected into either gift cards, exercise classes or goods. This progression system can not only motivate users but also encourage them to be more active.

* We see that when people are less active, they tend to have lesser sleep. Therefore, to improve one's quality sleep, a feature that can be implemented into the app is that users can enter their desired sleep time and awake time; then the app will calculate the hours and recommend a sleep time that is suitable depending on their sleep schedule. An app notification can also be sent out whenever no activity has been detected for a period of time that is considered sedentary. 

* For people who are or might be trying to lose weight, articles such as healthy food recipes and exercise videos can be recommended to users based on their preferences.

* For users that are active on Saturday, recommendations can be sent to them on other workouts or different intensities workout or jogs. For those who are least active especially on Sunday, Bellabeat can recommend challenging workout or activity and giving double points to targeted people to encourage them to be more active. 

